---
title: "WDA Student Workshop 2025: Part 1 R Basics"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    theme: darkly
    output_dir: "C:/Users/angel/OneDrive - The Pennsylvania State University/Documents/Teaching/R/wda-workshop-2025"
---

```{r setup, include=FALSE, cache = FALSE}


knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(),
fig.width = 12,
fig.height = 10,
fig.align = 'center')


```

### Outline

The sections of this workshop R script/R markdown are as follows:

 1. **Import Data**

 2. **Data Inspection**

 3. **Data Subsetting**

 4. **Using Functions and Dealing with NAs**

 5. **Working with Logical Operators**


### Data

 We are working with the species occurrence dataset from Global Biodiversity Facility Information (GBIF).

 - *sp* - a dataframe containing species occurrence information from aggregated sources (raw data is stored as a TSV or tab separated value)

### Import Data

```{r}

read.delim("data_raw/dicamptodon_tenebrosus_gbif/dicamptodon_tenebrosus_gbif.csv",sep = "\t", stringsAsFactors = T)

sp<-read.delim("data_raw/dicamptodon_tenebrosus_gbif/dicamptodon_tenebrosus_gbif.csv",sep = "\t", stringsAsFactors = T)


```


### Data Inspection

Here we learn how to inspect our data through the str, head, and names functions.

```{r}

str(sp)
head(sp)
names(sp)

```

### Data Subsetting

Here we learn how to subset our data by rows and columns.

```{r}


sp[1:6, ]

sp["eventDate"]

sp_date<-sp["eventDate"]


sp$eventDate

sp_date_vec<-sp$eventDate


sp[c("eventDate","institutionCode", "basisOfRecord")]

```

### Using Functions and Dealing with NAs

Here we learn how to use functions and learn to understand the impact of NAs on some functions.
```{r}


mean(sp$coordinateUncertaintyInMeters)


mean(sp$coordinateUncertaintyInMeters, na.rm = TRUE)


round(mean(sp$coordinateUncertaintyInMeters, na.rm = TRUE))

avg<-mean(sp$coordinateUncertaintyInMeters, na.rm = TRUE)
round(avg)



```

### Working with Logical Operators

Here we become familar with how logical operators work and how we can use these to manipulate our data.

```{r}
sp_inat <- sp[sp$institutionCode=="iNaturalist", ]



sp_inat_cas <- sp[sp$institutionCode=="iNaturalist" | sp$institutionCode=="CAS", ]

```


### Using Functions from Packages and Removing columns

Here we learn how to remove columns without a function and with a function.

```{r}

library(tidyverse)

sp_less <- sp[,-c(7,8)]



sp_less_tidy <- select(sp, -recordNumber)


```

